\documentclass{article}
\usepackage{alltt,multicol,array,tabularx}
\pagestyle{empty}
\usepackage{vpage}
\setpapersize{USletter}
\setmarginsrb{0.75in}{0.75in}{0.75in}{0.75in}{0pt}{0pt}{0pt}{0pt}


\newenvironment{mytt}{\alltt\def\\{\(\backslash\)}}{\endalltt}
\def\.{\char`\\}
\def\{{\char`\{}
\def\}{\char`\}}

\setcounter{secnumdepth}{0}


\newcommand\code[1]{$\langle$\emph{#1}$\rangle$}

\begin{document}
\raggedright

\parskip=0.5\baselineskip plus 0.2\baselineskip
\parindent=0pt
 
\hbox{}
\kern-1in
\advance\voffset by0.5in

\begin{center}
{\Large A One-Page Guide to Using \texttt{noweb} with {\LaTeX}}

\smallskip

{Norman Ramsey\\after Lee Wittenberg and Carl Gregory}
\end{center}


\begin{multicols}{2}[\section{The \texttt{noweb} file \texttt{foo.nw}}]

\vbox{
\centerline{\large\textbf{Documentation}}

\verb+@+ in column~1 of a line by itself starts a documentation chunk.

Text is in {\LaTeX}; use \verb+[[+\ldots \verb+]]+ for embedded code
fragments.

\texttt{noweave} automatically wraps file in
\begin{quote}
{\ttfamily\begin{tabular}{|l|}\hline
\.documentclass\{article\}\\
\.usepackage\{noweb\}\\
\.begin\{document\}\\
~~\vdots\\
\.end\{document\}\\
\hline
\end{tabular}}
\end{quote}
To avoid wrapper, put \verb+\documentclass+ and
\verb+\usepackage+ in your source and use \texttt{noweave~-delay}.
}

\vskip 0pt plus1in
\penalty-500
	
\vbox{
\centerline{\large\textbf{Code}}

\verb+<<+\ldots \emph{name of chunk}\ldots \verb+>>=+ in column~1 of a
line by itself starts a code chunk.

Code is in any language; use \verb+<<+\ldots \verb+>>+ to include
other code chunks by reference.

\emph{You can write multiple code chunks with the same name}; they are
concatenated.

Name chunks inside functions with verbs, e.g., 
\code{copy all lines from \texttt{\upshape in} to \texttt{\upshape
out}, adding markup}. 

Name chunks holding declarations with nouns, e.g., \code{private
function prototypes} or \code{definitions of exported functions}.

Name top-level ``root'' chunks after the files they are to 
represent, e.g., \verb+<<foo.c>>=+ or \verb+<<foo.h>>=+.
}
\vskip 0pt plus3in

\end{multicols}


\emph{For emacs users}: 
A first line of \verb+% -*- mode: Noweb; noweb-code-mode: c-mode -*-+
is useful (to set editing mode for code chunks).
The Unix \texttt{teTeX} distribution comes with helpful Info files
that cover {\LaTeX} and \texttt{dvips}.




\section{Typical Commands}

\noindent
\begin{tabularx}\textwidth{@{}>{\ttfamily}l>{\raggedright\arraybackslash}X}
notangle -L -Rfoo.c foo.nw > foo.c&Extract file \texttt{foo.c}, with \texttt{\#line}\\
notangle -Rfoo.h foo.nw | cpif foo.h&Extract file \texttt{foo.h} if it
has changed (without \texttt{\#line}).\\
\small noweave -filter l2h -index -html foo.nw > foo.html&
  Create HTML, converting documentation from {\LaTeX}.\\
noweave -x foo.nw > foo.tex& Create {\LaTeX} document with
cross-reference, no index.\\
latex foo; latex foo& Process {\LaTeX} into ``device-independent''
(\texttt{dvi}) file.\\
latex '\.scrollmode \.input foo'& Alternate {\LaTeX} command for use in
Makefile.\\
dvips -o foo.ps foo&Convert \texttt{foo.dvi} into PostScript.\\
\end{tabularx}


\begin{multicols}{2}[\section{Minimal {\LaTeX}}]

\emph{Special characters}\hfil\break
Not to be typed in text; OK in code\hfil\break
\hbox{\qquad \verb=% \ { } $ & # ^ _ ~=}

\emph{Escapes for specials}\hfil\break
\hbox{\qquad \verb=\% $\backslash$ \{ \} \$ \& \# \_=}

\emph{Blank line} is new paragraph.

Write \verb+``quotes''+ to get ``quotes.''  Never \verb+"quotes"+.

\newenvironment{sample}[1]{\begin{tabular}{@{}l>{\ttfamily}l}\emph{#1}}{\end{tabular}}

\begin{sample}{Titles}
&\.title\{\ldots\}\\
&\.author\{\ldots\}\\
&\.maketitle\\
\end{sample}



\begin{sample}{Sectioning}
&\.section\{\ldots\}\\
&\.subsection\{\ldots\}\\
&\.subsubsection\{\ldots\}\\
\end{sample}

\begin{sample}{Fonts}
&\.emph\{{\normalfont\emph{emphasis}}\}\\
&\.textbf\{{\normalfont\textbf{bold}}\}\\
&\.texttt\{{\normalfont\texttt{typewriter}}\}\\
&\.textit\{{\normalfont\textit{italics}}\}\\
&\.textsf\{{\normalfont\textsf{sans serif}}\}\\
&\.textsc\{{\normalfont\textsc{small caps}}\}\\
\end{sample}

\newdimen\onecol
\onecol=0.25\textwidth
\advance\onecol by -0.33in


{\noindent\parskip=0pt
\begin{tabular}{@{}lp{\onecol}p{\onecol}}
\emph{Lists}&
\texttt{List of items}
\par\kern-0.8\baselineskip
\begin{verbatim}
\begin{itemize}
\item one
\item two 
\item three
\end{itemize}
\end{verbatim}
\par\kern-2.5\baselineskip\hbox{}
&  List of items
\begin{itemize}
\item one
\item two 
\item three
\end{itemize}
\par\kern-2.5\baselineskip\hbox{}
\\
&\multicolumn2{l}{Also \texttt{\.begin\{enumerate\}} for numbered lists.}
\end{tabular}}

Get the right version; you want {\LaTeX}2$\varepsilon$, not {\LaTeX}~2.09.


\end{multicols}


\end{document}



